---
title: "Project"
output: html_document
date: "2025-02-12"
---
Jasper Dong and Tosca Wong

Question: What is the correlation between the number of businesses and Muni stops in San Francisco?

San Francisco is a highly concentrated city with a population of 808,988 people. This city, constructed on a 46.9 square mile area of land, is notorious for its constantly congested highways and traffic. Therefore, instead of driving, people often use public transportation to commute from one place to another. The San Francisco Municipal Railway (Muni) is a widely popular mode of public transportation in San Francisco. Muni is run by the San Francisco Municipal Transportation Agency and is composed of light rail, cable cars, buses, and streetcars. 

The overarching goal of our project is to analyze the relationship between presence of San Francisco Municipal Railway (MUNI) and location of businesses throughout San Francisco. We want to see where relationships between economic growth/decline and MUNI stops exist, and which neighborhoods this relationship is the most prominent in. According to the 2022 Muni Rider Survey, one third of Muni riders use Muni to travel to and from various businesses. Increased traffic in certain neighborhoods due to increases in business, could result in the need for more Muni stops in neighborhoods to relieve traffic congestion. According to the Transit Economic Benefits Study, the highest amount of Muni ridership occurs in neighborhoods of high congestion. We want to see which neighborhoods have the most Muni stops and analyze them for any correlation with the number of businesses established in the neighborhood.

***********Description of the dataset and exploratory data analysis (variable distribution, correlation etc),
(2 different plots and/or tables plus 3-4 paragraphs)

```{r}
## Load packages
library(terra)
library(tidyverse)
```

```{r}
## Unzipping files
unzip("Data.zip")
dir.create("muni", showWarnings = FALSE)
dir.create("nhood", showWarnings = FALSE)
unzip("MUNI.zip", exdir="muni")
unzip("nhood.zip", exdir="nhood")
```


```{r}
##Transforming longitude and latitude data so that it can be read as a vector 
##for the businesses data.

bis <- read.csv("Registered_Business_Locations_-_San_Francisco_20250212.csv", sep = ";")
regex1 <- "POINT \\("
bis$Business.Location <- gsub(regex1, "", bis$Business.Location)

regex2 <- "\\)"
bis$Business.Location <- gsub(regex2, "", bis$Business.Location)

bis_lonlat <- bis %>%
  separate(Business.Location, c("lon", "lat"), sep = " ") %>%
  mutate(lon = as.numeric(lon),
         lat = as.numeric(lat))

bisv <- vect(bis_lonlat, geom = c("lon", "lat"))
```

```{r}
## Vectorize Muni data
muni<- vect("muni/geo_export_7a61b6cd-add6-411a-bdc0-6a4be0880140.dbf")
muni

## Vectorize San Fransico Neighborhood Data
newv <- vect("nhood/geo_export_188f3512-6e48-45bc-ad4c-ef55315e7720.dbf")
newv

## Counting the amount of businesses in each neighborhood
count <- terra::extract(newv, bisv)

noNAcount <- count %>%
  filter(!is.na(nhood))

neighborhood <- noNAcount%>%
  group_by(nhood) %>%
  summarize(point_count = n())

join <- merge(newv, neighborhood, by="nhood")
point_count <- join[["point_count"]]

## Counting the amount of Muni stops in each neighborhood
muni_count <- terra::extract(newv, muni)
munistops <- muni_count%>%
  group_by(nhood)%>%
  summarize(munipoint_count = n())

muni_join <- merge(newv, munistops, by = "nhood")

busmuni_join <- merge(muni_join, join, by = "nhood")

```

```{r, fig.height= 16, fig.width= 16}
## Plot Muni and Business data

par(mfrow = c(1,2))

##Picking color palette for Businesses map
YlOrRd <- RColorBrewer::brewer.pal(n=9, name ="YlOrRd")
colors = colorRampPalette(c("#FFEDA0", "#BD0026"))

##Picking color palette for Businesses map
Purple <- RColorBrewer::brewer.pal(n=9, name ="Purples")
colors1 = colorRampPalette(c("#EFEDF5", "#54278F"))

## Businesses in Each Neighborhood Map
plot(join, "point_count", breaks= 10, col = colors(10), background = "lightblue", 
     main= "Businesses in SF Neighborhoods", axes=FALSE, cex.main = 1.7)
text(x=-122.435, y=37.7072, "ABT 181N Jasper Dong & Tosca Wong", col = "black", cex=0.25)
lines(newv, col="black")
north(xy= "topleft", type=2, cex= 1.2)
sbar(5, xy=c(-122.5,37.824), type="line", divs=2, labels=c(0,500,1000), 
   scaleby=100, lonlat=NULL, cex=0.8, below = "km")

## Muni Stops in Each Neighborhood Map
plot(muni_join, "munipoint_count", breaks = 10, col = colors1(10), background = "lightblue", 
     main= "Muni Stops in SF Neighborhoods", axes=FALSE, cex.main = 1.7)
text(x=-122.435, y=37.7072, "ABT 181N Jasper Dong & Tosca Wong", col = "black", cex=0.25)
lines(newv, col="black")
north(xy= "topleft", type=2, cex= 1.2)
sbar(5, xy=c(-122.5,37.824), type="line", divs=2, labels=c(0,500,1000), 
   scaleby=100, lonlat=NULL, cex=0.8, below = "km")
```
The first map visualizes the number of businesses in each San Francisco neighborhood. Neighborhoods with fewer businesses are represented in yellow, while higher numbers of businesses are represented in red. There is a high concentration of businesses in the northwest quadrant of San Francisco.

The second map visualizes the number of Muni stops in each San Francisco neighborhood. Neighborhoods with fewer businesses are represented in light purple, while higher numbers of businesses are represented in dark purple. There is a high concentration of Muni Stops in the southwest quadrant of San Francisco.

```{r fig.height= 4, fig.width= 16}

par(mfrow = c(1,2))

##Bar Plot indicating the number of businesses in each neighborhood
barplot(height = neighborhood$point_count, names.arg = neighborhood$nhood,
  main = "Number of Businesses in Each SF Neighborhood",
  xlab = "Neighborhood", ylab = "Number of Businesses",
  col = "navy", cex.names =0.35, las = 2, cex.axis = 0.7)

##Bar Plot indicating the number of Muni stops in each neighborhood
barplot(height = munistops$munipoint_count, names.arg = munistops$nhood,
  main = "Number of Muni Stops in Each SF Neighborhood",
  xlab = "Neighborhood", ylab = "Number of Muni Stops",
  col = "navy", cex.names =0.35, las = 2, cex.axis = 0.7)
```
The first plot illustrates the relationship between the number of businesses and neighborhoods in San Francisco. The majority of businesses are located in the financial district as it is the center of finance and corporate headquarters. This district was established as a financial center during the Gold Rush and continues to expand today with the rise in countless skyscrapers to increase the density of businesses in the area. 

The second plot illustrates the relationship between the number of Muni stops and neighborhoods in San Francisco. The least amount of Muni stops are located in the Lincoln Park and Sea Cliff District. The lack of Muni stops in the Lincoln Park district may be because a majority of the land is dedicated for parkland and recreational use. Therefore, there is a lack of roads and Muni stops located in this district. Sea Cliff also has minimal Muni stops as this district is smaller in area and known for its elaborate homes. Most of the people who reside in this district have their own vehicles, reducing their reliance on public transportation. On the other hand, Bayview has the most Muni stops. This may be because it is one of the largest districts in area, but it is historically an underserved area. The community there may lack the resources to have personal transportation, and therefore are heavily reliant on public transportation. Due to the high demand in Bayview, there are more Muni stops to improve connectivity. 

```{r}
##Number of Muni Stops vs Number of Businesses in San Francisco
plot(busmuni_join$munipoint_count, busmuni_join$point_count,
    main = "Number of Muni Stops vs Number of Businesses in San Francisco", 
    cex.main = 1.2,
    xlab = "Number of Muni Stops", ylab = "Number of Businesses",
    pch=16, cex=1, col='blue'
    )
abline(a= busmuni_join$point_count, b = busmuni_join$munipoint_count, col='red',
       lwd = 1)
```
This scatterplot depicts the relationship between the number of Muni stops and the number of businesses. The linear regression line shows a slight positive correlation between Muni stops and businesses. Outliers are present and may have affected the regression line, causing our two variables to be less correlated than they should have been.

```{r}
## Checking for outliers
par(mfrow=c(1, 2))
boxplot(busmuni_join$munipoint_count, main="Number of MUNI Stops")
boxplot(busmuni_join$point_count, main="Number of Businesses")
```

```{r}
# Obtain values of outliers
muniout <- boxplot.stats(busmuni_join$munipoint_count)$out
busout <- boxplot.stats(busmuni_join$point_count)$out
# Obtaining indices
muniout_idx <- which(busmuni_join$munipoint_count %in% muniout)
busout_idx <- which(busmuni_join$point_count %in% busout)
outidx <- c(muniout_idx, busout_idx)
# Removing outliers
df <- values(busmuni_join)
dfclean <- df[-outidx, ]
# Re-plotting boxplots
par(mfrow=c(1, 2))
boxplot(dfclean$munipoint_count, main="Number of MUNI Stops")
boxplot(dfclean$point_count, main="Number of Businesses")
```

```{r}
# Fitting linear regression model
linearMod <- lm(point_count ~ munipoint_count, data=dfclean) 
summary(linearMod)
```

********Discussion




```{r}
##### RUN THIS BEFORE PUSHING TO GITHUB
##### TO AVOID ACCIDENTALLY SENDING BIG FILES
file.remove("Registered_Business_Locations_-_San_Francisco_20250212.csv")
unlink("muni", recursive = T)
unlink("nhood", recursive = T)
```

*********Conclusion

References
Brophy, Jeanne. “2022 Muni Rider Survey.” SFMTA, 6 Feb. 2023, www.sfmta.com/blog/2022-muni-rider-survey. 

SFMTA Citizens Advisory Committee2. “SFMTA.” Transit Economic Benefits Study, 3 Sept. 2015, www.sfmta.com/media/29156/download?inline=.


