---
title: "Project"
output: html_document
date: "2025-02-12"
---
Jasper Dong and Tosca Wong

Question: What is the correlation between the number of businesses and Muni stops in San Francisco?

San Francisco is a highly concentrated city with a population of 808,988 people. This city, constructed on a 46.9 square mile area of land, is notorious for its constantly congested highways and traffic. Therefore, instead of driving, people often use public transportation to commute from one place to another. The San Francisco Municipal Railway (Muni) is a widely popular mode of public transportation in San Francisco. Muni is run by the San Francisco Municipal Transportation Agency and is composed of light rail, cable cars, buses, and streetcars. 

The overarching goal of our project is to analyze the relationship between presence of San Francisco Municipal Railway (MUNI) and location of businesses throughout San Francisco. We want to see where relationships between economic growth/decline and MUNI stops exist, and which neighborhoods this relationship is the most prominent in. According to the 2022 Muni Rider Survey, one third of Muni riders use Muni to travel to and from various businesses. Increased traffic in certain neighborhoods due to increases in business, could result in the need for more Muni stops in neighborhoods to relieve traffic congestion. According to the Transit Economic Benefits Study, the highest amount of Muni ridership occurs in neighborhoods of high congestion. We want to see which neighborhoods have the most Muni stops and analyze them for any correlation with the number of businesses established in the neighborhood.

***********Description of the dataset and exploratory data analysis (variable distribution, correlation etc),
(2 different plots and/or tables plus 3-4 paragraphs)

```{r}
## Load packages
library(terra)
library(tidyverse)
library(ggplot2)
```

```{r}
## Unzipping files
unzip("Data.zip")
dir.create("muni", showWarnings = FALSE)
dir.create("nhood", showWarnings = FALSE)
unzip("MUNI.zip", exdir="muni")
unzip("nhood.zip", exdir="nhood")
```


```{r}
##Transforming longitude and latitude data so that it can be read as a vector 
##for the businesses data.

bis <- read.csv("Registered_Business_Locations_-_San_Francisco_20250212.csv", sep = ";")
regex1 <- "POINT \\("
bis$Business.Location <- gsub(regex1, "", bis$Business.Location)

regex2 <- "\\)"
bis$Business.Location <- gsub(regex2, "", bis$Business.Location)

bis_lonlat <- bis %>%
  separate(Business.Location, c("lon", "lat"), sep = " ") %>%
  mutate(lon = as.numeric(lon),
         lat = as.numeric(lat))

bisv <- vect(bis_lonlat, geom = c("lon", "lat"))
```

```{r}
## Vectorize Muni data
muni<- vect("muni/geo_export_7a61b6cd-add6-411a-bdc0-6a4be0880140.dbf")
muni

## Vectorize San Fransico Neighborhood Data
newv <- vect("nhood/geo_export_188f3512-6e48-45bc-ad4c-ef55315e7720.dbf")
newv

## Counting the amount of businesses in each neighborhood (**the line below also takes at least 10 mins to run)
count <- terra::extract(newv, bisv)

noNAcount <- count %>%
  filter(!is.na(nhood))

neighborhood <- noNAcount%>%
  group_by(nhood) %>%
  summarize(point_count = n())

join <- merge(newv, neighborhood, by="nhood")
point_count <- join[["point_count"]]

## Counting the amount of Muni stops in each neighborhood
muni_count <- terra::extract(newv, muni)
munistops <- muni_count%>%
  group_by(nhood)%>%
  summarize(munipoint_count = n())

muni_join <- merge(newv, munistops, by = "nhood")

busmuni_join <- merge(muni_join, join, by = "nhood")
```
```{r}
# Converting count data to proportion data
join$prop <- join$point_count/sum(join$point_count)

muni_join$prop <- muni_join$munipoint_count/sum(muni_join$munipoint_count)

join
muni_join
```


Sources
DataSF MUNI Transit Stop Data provided a shapefile with location data for all Muni Stops.

DataSF Registered Business Locations provided a CSV with the location for all tax-paying businesses in San Francisco. According to DataSF, the data is collected through business registration applications and taxpayer filings.
 
DataSF Analysis Neighborhood provided a shapefile with census data about each of the 41 neighborhoods in San Francisco. 


```{r, fig.height= 16, fig.width= 16}
## Plot Muni and Business data

par(mfrow = c(1,2))

##Picking color palette for Businesses map
YlOrRd <- RColorBrewer::brewer.pal(n=9, name ="YlOrRd")
colors = colorRampPalette(c("#FFEDA0", "#BD0026"))

##Picking color palette for Businesses map
Purple <- RColorBrewer::brewer.pal(n=9, name ="Purples")
colors1 = colorRampPalette(c("#EFEDF5", "#54278F"))

## Businesses in Each Neighborhood Map
plot(join, "prop", breaks= 10, col = colors(10), background = "lightblue", 
     main= "Businesses in SF Neighborhoods", axes=FALSE, cex.main = 1.7)
text(x=-122.435, y=37.7072, "ABT 181N Jasper Dong & Tosca Wong", col = "black", cex=0.75)
lines(newv, col="black")
north(xy= "topleft", type=2, cex= 1.2)
sbar(5, xy=c(-122.5,37.824), type="line", divs=2, labels=c(0,500,1000), 
   scaleby=100, lonlat=NULL, cex=0.8, below = "km")

## Muni Stops in Each Neighborhood Map
plot(muni_join, "prop", breaks = 10, col = colors1(10), background = "lightblue", 
     main= "Muni Stops in SF Neighborhoods", axes=FALSE, cex.main = 1.7)
text(x=-122.435, y=37.7072, "ABT 181N Jasper Dong & Tosca Wong", col = "black", cex=0.75)
lines(newv, col="black")
north(xy= "topleft", type=2, cex= 1.2)
sbar(5, xy=c(-122.5,37.824), type="line", divs=2, labels=c(0,500,1000), 
   scaleby=100, lonlat=NULL, cex=0.8, below = "km")
```
The first map visualizes the number of businesses in each San Francisco neighborhood. Neighborhoods with fewer businesses are represented in yellow, while higher numbers of businesses are represented in red. There is a high concentration of businesses in the northwest quadrant of San Francisco.

The second map visualizes the number of Muni stops in each San Francisco neighborhood. Neighborhoods with fewer businesses are represented in light purple, while higher numbers of businesses are represented in dark purple. There is a high concentration of Muni Stops in the southwest quadrant of San Francisco.

```{r fig.height= 6, fig.width= 16}
par(mfrow = c(1,2))

##Bar Plot indicating the proportion of businesses to the total number of businesses in each SF neighborhood
barplot(height = neighborhood$point_count/sum(neighborhood$point_count), names.arg = neighborhood$nhood,
  main = "Proportion of Businesses in Each SF Neighborhood",
  xlab = "Neighborhood", ylab = "Businesses",
  col = "navy", cex.names =0.4, las = 2, cex.axis = 0.7)

##Bar Plot indicating the proportion of Muni stops to the total number of Muni stops in each neighborhood
barplot(height = munistops$munipoint_count/sum(munistops$munipoint_count), names.arg = munistops$nhood,
  main = "Proportion of Muni Stops in Each SF Neighborhood",
  xlab = "Neighborhood", ylab = "Muni Stops",
  col = "navy", cex.names =0.4, las = 2, cex.axis = 0.7)
```
The first plot illustrates the relationship between the proportion of businesses to the total businesses and neighborhoods in San Francisco. The majority of businesses are located in the financial district as it is the center of finance and corporate headquarters. This district was established as a financial center during the Gold Rush and continues to expand today with the rise in countless skyscrapers to increase the density of businesses in the area. 

The second plot illustrates the relationship between the proportion of Muni stops to the total businesses and neighborhoods in San Francisco. The least amount of Muni stops are located in the Lincoln Park and Sea Cliff District. The lack of Muni stops in the Lincoln Park district may be because a majority of the land is dedicated for parkland and recreational use. Therefore, there is a lack of roads and Muni stops located in this district. Sea Cliff also has minimal Muni stops as this district is smaller in area and known for its elaborate homes. Most of the people who reside in this district have their own vehicles, reducing their reliance on public transportation. On the other hand, Bayview has the most Muni stops. This may be because it is one of the largest districts in area, but it is historically an underserved area. The community there may lack the resources to have personal transportation, and therefore are heavily reliant on public transportation. Due to the high demand in Bayview, there are more Muni stops to improve connectivity. 


```{r, fig.height=10, fig.width=6}
## Checking for outliers
par(mfrow=c(1, 2))
boxplot(muni_join$prop, main="Proportion of MUNI Stops", 
        outcex = 2.5, outpch = 4)
boxplot(join$prop, main="Proportion of Businesses", 
        outcex = 2.5, outpch = 4)
```

```{r, fig.height=10, fig.width=6}
busmuni_join$muniprop <- busmuni_join$munipoint_count/sum(busmuni_join$munipoint_count)
busmuni_join$pointprop <- busmuni_join$point_count/sum(busmuni_join$point_count)
# Obtain values of outliers
muniout <- boxplot.stats(busmuni_join$muniprop)$out
busout <- boxplot.stats(busmuni_join$pointprop)$out
# Obtaining indices
muniout_idx <- which(busmuni_join$muniprop %in% muniout)
busout_idx <- which(busmuni_join$pointprop %in% busout)
outidx <- c(muniout_idx, busout_idx)
# Removing outliers
df <- values(busmuni_join)
dfclean <- df[-outidx, ]
# Re-plotting boxplots
par(mfrow=c(1, 2))
boxplot(dfclean$muniprop, main="Proportion of MUNI Stops")
boxplot(dfclean$pointprop, main="Proportion of Businesses")
```

```{r}
# Fitting linear regression model
lin <- lm(pointprop ~ muniprop, data=dfclean) 
linsum <- summary(lin)
linsum
```
```{r}
# Obtaining correlation between MUNI Stops and Businesses
correlation = cor(dfclean$muniprop, dfclean$pointprop)
correlation
```
```{r}
# Plotting linear relationship between MUNI stops and Businesses
ggplot(data = dfclean, aes(x = muniprop, y = pointprop)) +
  geom_point() +
  geom_smooth(method='lm', se=FALSE, color = "steelblue") +
  theme_minimal() +
  labs(x = 'Proportion of MUNI Stops', y = 'Proportion of Businesses',
       title = 'Linear Regression of MUNI Stops vs. Businesses') +
  theme(plot.title = element_text(hjust=0.5, size=12, face='bold')) +
  geom_text(x = 0.0075, y = 0.03, 
            label = paste0("R^2 = ", round(linsum$r.squared, 4))) +
  geom_text(x = 0.0083, y = 0.028, 
            label = paste0("r = ", round(correlation, 4)))
```
```{r, message = F}
# Reading in demographic data
demo <- read_csv("Demo.csv")
# Filtering by race and time categories
demo <- demo[which(demo$demographic_category == "race"),]
demo <- demo[which(demo$end_year == 2023),]
# Selecting variables of interest
democlean <- demo |> 
  select(all_of(c("geography_name",
                  "demographic_category_label", 
                  "estimate")))
colnames(democlean)[1] <- "nhood"
# Obtaining unique race groups
racelab <- sort(unique(democlean$demographic_category_label))
racelab
```


```{r}
# Splitting data frame into separate race groups
splitrace <- split(democlean, democlean$demographic_category_label)
# Dropping unnecessary categories
splitrace <- splitrace[-c(6, 7)] 
# Obtaining estimate counts
## Selecting count data and coercing into data frame
racecount <- lapply(splitrace, function(x){x |> select("estimate")})
names(racecount) <- c("native", "asian", "black", "hisp", "pi", "white")
race <- as.data.frame(mapply(cbind, racecount))
## Calculating total population counts for each nhood
race$total <- rowSums(race)
race$nhood <- splitrace[[1]]$nhood
## Turning counts into proportions by each nhood
race[,1:6] = round(race[,1:6]/race$total, 4)
colnames(race)[1:6] <- names(racecount)
head(race)
```

```{r}
# Merging into SpatVector
demo.join <- merge(busmuni_join, race, by = "nhood")
demo.join
```

```{r}
# Turning into data frame
dfrace = values(demo.join)
# Removing previous outliers
outidxcomb <- c()
for(i in 4:11){
  out <- boxplot.stats(dfrace[,i])$out
  outidx <- which(dfrace[,i] %in% out)
  outidxcomb <- c(outidxcomb, outidx)
}

dfrace.clean <- dfrace[-unique(outidxcomb), ]
head(dfrace.clean)
```

```{r}
library(randomForest)
modelRF <- randomForest(pointprop ~ muniprop + native + asian + black +
                 hisp + pi + white, data = dfrace.clean)

modelRF
```


********Discussion
```{r}
##### RUN THIS BEFORE PUSHING TO GITHUB
##### TO AVOID ACCIDENTALLY SENDING BIG FILES
file.remove("Registered_Business_Locations_-_San_Francisco_20250212.csv")
unlink("muni", recursive = T)
unlink("nhood", recursive = T)
```

Conclusion
We found that there is a slight positive correlation between the amount of Muni stops and the number of business in each San Francisco neighborhood. There were some outliers such as the Financial district and Bayview Hunters Point that caused our result to become skewed. Removal of the outliers helped us to gain a more accurate representation of the correlation between our two variables. Possible next steps include analyzing ridership and zoning data along with the data found in this project to see how people move throughout the City of San Francisco during different times of the day and week. This could help determine where and when people travel to areas for activities such as work or entertainment, aiding in a more focused and intentional plan for growth.

References
Brophy, Jeanne. “2022 Muni Rider Survey.” SFMTA, 6 Feb. 2023, www.sfmta.com/blog/2022-muni-rider-survey. 

SFMTA Citizens Advisory Committee2. “SFMTA.” Transit Economic Benefits Study, 3 Sept. 2015, www.sfmta.com/media/29156/download?inline=.


